\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-30 coqimpl.tex

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We support ``extensible'' inductive types from
\ref{chg:extensible-inductive-type} in Coq to some extent.\footnote{Because to give a satisfactory formal definition and semantic for extensible inductive type is not something this paper aims at, and obviously requires more theoretical research effort}\YZ{I do consider that your plugin supports extensible inductive types. It's just that they are implemented as a plugin, rather than in the Coq core.}\EDJreply{I rephrased it. It is more like I don't think this is really the final form of extensible inductive type that everyone is happy about}

@@

To construct a recursor as in \ref{chg:extensible-inductive-type}, we need to create for each
constructor a handler (function) and aggregate them into a family, and thus that family (e.g. "neg_handler" in \cref{fig:plugin-example2}) can be considered as a large matching-case statement. We use "FRecursor" to declare a recursor (e.g. "neg") by specifying the motive, the aiming inductive type and the handler family. Once recursor is created it will look like any other function fields (e.g. "neg"). Here in \cref{fig:plugin-example2} we apply "neg" to "tt" as an exemplar field. The inheritance of recursor is mainly delegated by the inheritance of the handler family---if the user extend with new constructor (e.g. "uu"), then the inherited family (e.g. "neg_handler") will need to extend correspondingly (e.g. "neg_handler.uu"). If not, the plugin will error with ``Non-exhaustiveness Pattern'' when inheriting the field "neg".

@@

To summarize, the key insight here is that, to make each recursor
handler inheritable, we need to seal an abstraction around the inductive
type (e.g., module "b_3"); but to construct a recursor (and carry out the
exhaustiveness checking), we need to break this abstraction and see the
concrete definition of the inductive type (e.g., module "b_rec_12"). Our
meta-theory need to handle these two seemingly contradicting ideas
simultaneously.

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is
not expressive enough to really fulfill \ref{chg:extensible-inductive-type}. For
example, we cannot prove the constructors are injective, i.e.,
\mintinline{Coq}{self_B.tt ≠ self_B.ff} because from the perspective of Coq they are just two terms declared in the module type.\YZ{Is it really impossible to prove constructor injectivity without partial recursors and with recursors only? My confusion after reading these paras is that I still don't know why partial recursors are needed.}\EDJreply{The issue is exactly, at that point of programming, for extensibility, there is no access to the concrete recursor. Concrete recursors also cannot be postulated as part of the interface because it breaks extensibility.}
In the original setting, to
prove injectivity of constructors of a (vanilla) inductive type, we only
require a recursor and its computational rules. However, as we pointed
out earlier, the recursor of an inductive type cannot be exposed into
the context for the sake of extensibility. Hence, we need to alter the
formulation of the recursor: we need to have a better
characterization of ``extensible inductive types'' exported in the
compiled module type as a part of the interface for the following fields
to use.

The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

What's more, partial recursor is theoretically enough for the
"injection" and "discriminate" tactics. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{self_B.tt ≠
self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

In fact, the above "reflect" function implies that propositional partial
recursor is good enough to some extent, at least for non-indexed
inductive type: because we can ``embed'' vanilla inductive types into
types supporting a partial recursor. Because "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), .., 
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of
  "C" and "prec" of "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered as a constructor, because they can be reflected
to real constructors of "C" using the left inverse "linv" (directly generalization of "reflect : self_B.b → option bool" of the above). Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete characterization of all the extension of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases the users just want global reasoning
instead of an extensible proof. For example, in STLC, we expect
``\mintinline{Coq}{~ value (tm_app x y)}'' to hold in all future
extensions of STLC---applications will never be considered as values
by an extension.
We know the correct way to prove the proposition
\mintinline{Coq}{value_not_app : value (tm_app x y) -> False.} is to use
"FRecursor" to inductively reason "value (tm_app x y)". However, this
means that every time we extend "value" with new value forms, we will
need to extend corresponding inductive cases for "app_not_value". We can
imagine the newly extended proof to be boring case analysis that the
newly added value forms are not "tm_app". What's worse, similar
scenario can happen on other statement like \mintinline{Coq}{~ value
(tm_if cond x y)}. Everything here is extended correctly but we end up
having boilerplate code, \textbf{semantically}.
\begin{figure}[!htb]
\begin{minted}{Coq}
Family STLC.
  FInductive tm : Set := ... 
  FInductive value : tm -> Prop := ... 
  Closing Fact value_not_app : forall x y, ~ value (tm_app x y) 
      by { intros x y H; inversion H; eauto }.
  Fail Ltac inv := 
    match goal with 
    | [h : value (tm_app _ _) |- ] => destruct (value_not_app _ _ h)
    (* Fail: value, tm_app, value_not_app unfound *)
    end. 
  MetaData tactic1.
    Ltac inv := (* ... the same definition as above ... *)
  EndMetaData.
EndFamily.
\end{minted}  
\caption{Example for Global Reasoning and MetaData}\label{fig:plugin-example-global-reasoning-meta-data}
\end{figure}

Our proposal is to allow global reasoning, which will look like a
constraint inside the family. We provide "Closing Fact" as a command to
carry out this global reasoning, attached with a proof script and proof
term. This "Closing Fact" will look like an assertion during the
construction of the enclosing family. This assertion will only be
verified when the whole family is closed and compiled into a module,
with the help of the attached proof script and proof terms.

Here we show the example to prove "value_not_app" in
\cref{fig:plugin-example-global-reasoning-meta-data}, the script inside
curly brackets after "by" is the attached proof script that will be
executed upon "EndFamily", which triggers the compilation of the family
being defined.
% \YZ{Please check if my wording is correct.}\EDJreply{Yes. Resolved}
Since we are dealing with a concrete inductive type during compilation,
we can use Coq's "inversion" tactic to complete this proof.
This "Closing Fact", as an assertion, will have no problem being
inherited, but the proof script will be executed every time a derived
family is compiled.

The downside is that "Closing Fact" cannot immediately check the correctness of the
attached proof term/proof script because this verification only happens when we
finished defining a family, and thus a bit harder to handle.\YZ{'harder' in what way? I suppose you mean the "Fail Ltac" in Figure 5?}\EDJreply{The proof script cannot be run immediately to reply the user whether the proof script is even correct. This is bad engineering to some extent. In those bad cases, because error only happens at the the very end ``EndFamily'' , the user will need to go back to the place of Closing Fact to adjust the proof script, or to adjust other parts. I am not sure if early check is possible to be done (engineering-wise) I think it is possible (imagine I just do an early compilation, i.e. only compile part of the family to the point closing fact is used) but it is a big engineering difficulty for me.}

Ultimately, "Closing Fact" is asserting a constraint upon all the
possible extension of the surrounding family and claiming a proof script
can solve the constraints.\YZ{This is a good one-liner that summarizes the intended usage scenarios of "Closing Fact". Should probably bring it out earlier.}
We expect it to be used only sporadically for maintainability.

@@

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family,
the family data is invisible to Coq internal and thus causes some

inconvenience. For example, in
\cref{fig:plugin-example-global-reasoning-meta-data} the first attempt
of defining tactic "inv" fails. Because the earlier defining "tm",
"value", and "value_not_app" are currently just data structure in the
plugin and not visible for Coq internal (only visible after the family
is compiled into module).

@@

>>>>

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We support extensible inductive types in Coq mentioned in
\ref{chg:extensible-inductive-type} but we will not be aiming at the research of its formal semantic.\footnote{Because to give a satisfactory formal definition and semantic for extensible inductive type is not something this paper aims at, and obviously requires more theoretical research effort}\YZ{I do consider that your plugin supports extensible inductive types. It's just that they are implemented as a plugin, rather than in the Coq core.}\EDJreply{I rephrased it. It is more like I don't think this is really the final form of extensible inductive type that everyone is happy about}

@@

Recall, to construct a recursor, we need to create for each
constructor a handler (function) and aggregate them into a family, and thus that family (e.g. "neg_handler" in \cref{fig:plugin-example2}) can be considered as a large matching-case statement. We use "FRecursor" to declare a recursor (e.g. "neg") by specifying the motive, the aiming inductive type and the handler family. Once recursor is created it will look like any other function fields (e.g. "neg"). Here in \cref{fig:plugin-example2} we apply "neg" to "tt" as an exemplar field. The inheritance of recursor is mainly delegated by the inheritance of the handler family---if the user extend with new constructor (e.g. "uu"), then the inherited family (e.g. "neg_handler") will need to extend correspondingly (e.g. "neg_handler.uu"). If not, the plugin will error with ``Non-exhaustiveness Pattern'' when inheriting the field "neg".

@@

To summarize, (1) to make each recursor
handler inheritable, we need to seal an abstraction around the inductive
type (e.g., module "b_3"); but to construct a recursor (and carry out the
exhaustiveness checking), we need to break this abstraction and see the
concrete definition of the inductive type (e.g., module "b_rec_12"). Our
meta-theory need to handle these two seemingly contradicting ideas
simultaneously. (2) \ref{chg:extensible-inductive-type} is resolved by this \textit{decoupling} of the syntax of implementation of recursor handlers and exhaustiveness checking. The former is handled by family inheritance and thus avoid the boilerplate code; and the non-inheritance of the compiled recursor ensures the exhaustiveness checking happens for every recursor.

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases an extensible proof brings more boilerplate code. For example, in STLC, we expect
``\mintinline{Coq}{~ value (tm_app x y)}'' to hold in all future
extensions of STLC---applications will never be considered as values
by any extension.
We know the correct way to prove the proposition
\mintinline{Coq}{value_not_app : value (tm_app x y) -> False.} is to use
"FRecursor" to inductively reason "value (tm_app x y)". However, this
means that every time we extend "value" with new value forms, we will
need to extend corresponding inductive cases for "app_not_value". We can
imagine the newly extended proof to be boring case analysis that the
newly added value forms are not "tm_app". What's worse, similar
scenario can happen on other statement like \mintinline{Coq}{~ value
(tm_if cond x y)}. Everything here is extended correctly but we end up
having boilerplate code, \textbf{semantically}.
\begin{figure}[!htb]
\begin{minted}{Coq}
Family STLC.
  FInductive tm : Set := ... 
  FInductive value : tm -> Prop := ... 
  Closing Fact value_not_app : forall x y, ~ value (tm_app x y) 
      by { intros x y H; inversion H; eauto }.
  Fail Ltac inv := 
    match goal with 
    | [h : value (tm_app _ _) |- ] => destruct (value_not_app _ _ h)
    (* Fail: value, tm_app, value_not_app unfound *)
    end. 
  MetaData tactic1.
    Ltac inv := (* ... the same definition as above ... *)
  EndMetaData.
EndFamily.
\end{minted}  
\caption{Example for Global Reasoning and MetaData}\label{fig:plugin-example-global-reasoning-meta-data}
\end{figure}

Our proposal is to allow global reasoning by providing "Closing Fact" command. Ultimately, "Closing Fact" is asserting a constraint upon all the
possible extension of the surrounding family and claiming that a proof script
can solve the constraints.\YZ{This is a good one-liner that summarizes the intended usage scenarios of "Closing Fact". Should probably bring it out earlier.}\EDJreply{I put it here as you suggested but I think this one-liner at here will cause more confusion because it is too abstract.}
The assertion will only be verified when the whole family is closed and compiled into a module.

% We provide "Closing Fact" as a command to
% carry out this global reasoning, attached with a proof script and proof
% term. This "Closing Fact" will look like an assertion during the
% construction of the enclosing family. This assertion will only be
% verified when the whole family is closed and compiled into a module,
% with the help of the attached proof script and proof terms.

Here we show the example to prove "value_not_app" in
\cref{fig:plugin-example-global-reasoning-meta-data}, the script inside
curly brackets after "by" is the attached proof script that will be
executed upon "EndFamily", which triggers the compilation of the family
being defined.
% \YZ{Please check if my wording is correct.}\EDJreply{Yes. Resolved}
Since we are dealing with a concrete (vanilla) inductive type during compilation,
we can use Coq's "inversion" tactic to complete this proof.
This "Closing Fact", as an assertion, will have no problem being
inherited, but the proof script will be executed every time a derived
family is compiled.

The downside is that "Closing Fact" cannot immediately check the correctness of the
attached proof term/proof script because this verification only happens when we
finished defining a family, and thus a bit harder to handle.\YZ{'harder' in what way? I suppose you mean the "Fail Ltac" in Figure 5?}\EDJreply{The proof script cannot be run immediately to reply the user whether the proof script is even correct. This is bad engineering to some extent. In those bad cases, because error only happens at the the very end ``EndFamily'' , the user will need to go back to the place of Closing Fact to adjust the proof script, or to adjust other parts. I am not sure if early check is possible to be done (engineering-wise) I think it is possible (imagine I just do an early compilation, i.e. only compile part of the family to the point closing fact is used) but it is a big engineering difficulty for me.}
We expect it to be used only sporadically for maintainability, and as an alternative to using "FRecursor". This is called global reasoning because this is a post-hoc reasoning when the family as a \textit{whole} is compiled into a module.

@@

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family,
the family data is invisible to Coq internal and thus causes some
inconvenience. For example, in
\cref{fig:plugin-example-global-reasoning-meta-data} the first attempt
of defining tactic "inv" fails. Because the earlier defining "tm",
"value", and "value_not_app" are currently just data structure in the
plugin and not visible for Coq internal (only visible after the family
is compiled into module).

@@

% Following needs a rewrite

\textbf{Propositional Partial Recursor.} Notice that, for both "FRecursor" and "Closing Fact", they are not necessarily ``\textit{conservative extension}''--they will possibly disrupt arbitrary extension of inductive type. For example, for \cref{fig:STLC-example}, once we proved \mintinline{Coq}{var_not_value : ∀ i, ~ value (tm_var i)} in the "Family STLC" either using "FRecursor" or "Closing", then any children family of "STLC" cannot extend a new constructor \mintinline{Coq}{vvar: ∀ i, value (tm_var i)}.

This raise an interesting question: what propositions are conservative extension? For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a ``best'' proposition (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

Since injectiveness of the constructors are derived by recursor in vanilla inductive type, an educated guess would be related to recursor. But we know the original recursor won't work because the mere existence of the original recursor will break extensibility. Thus we need to modify it.
The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

The first thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{self_B.tt ≠
self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

Even though we cannot prove propositional partial recursor is the best,
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (with certain constraints): because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$ s.t.

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), .., 
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of
  "C" and "prec" of "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
The reason we have this weird looking \ref{prop:detectable-partiality} is because the construction of "linv"---at one point, using "prec" to construct "linv", we need to prove \mintinline{Coq}{∀ a, (Bᵢ a → option C) → option C}, generally not provable, only if we have \mintinline{Coq}{option (Bᵢ a → C)} as an argument instead. This \ref{prop:detectable-partiality} is exactly transforming former to the later: recall that, according to the semantic of inductive type, \mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using "linv" on the substructure of "D". This function points out each of the translation result from "D" to "C" and some of them are "None" because of failing. For us, we only care about if \textbf{any} of the substructure of "D" is failed to translate---if any one substructure fails, then the whole transltion fails and thus we should have "None" as return.

\ref{prop:detectable-partiality} is derivable for all the finitary branching inductive type\footnote{By simply enumerating through the finite domain "Bᵢ a"}---that means all of the examples shown in this paper. \ref{prop:detectable-partiality} is not supported when using function (infinitary branching) in the constructors, for example, the case of using HOAS to represent functions in "tm". In those cases, we still have the fact that partial recursor is a conservative extension, but we cannot show that partial recursor is the extensional characterization any more。

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered as a constructor, because they can be reflected
to real constructors of "C" using the left inverse "linv" (directly generalization of "reflect : self_B.b → option bool" of the above). Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete extensional characterization of all the extension of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}
